---
title: "My alien analysis"
author: "Ian Handel"
output: html_document
---

# Set up RStudio
See handout...

1. Folder
2. Copy data
3. Setup project
4. Set working directory
5. Turn off inline RMarkdown

# Load packages

```{r, warn = FALSE}

library(tidyverse)
library(readxl)
library(skimr)
library(knitr)

```
# Data cleaning 

Import data
```{r}
alien <- read_excel("aliendata_20180331.xlsx", sheet = 1, na = "-999")
```

Fill down blank regions
```{r}
alien <- alien %>%
  fill(region)
```

Use select to get rid of name column.
```{r}
alien <- alien %>%
  select(-name)
```

Tidy ID column
```{r}
alien <- alien %>%
  mutate(subject = str_pad(subject, width = 3, side = "left", pad = "0")) %>% 
  mutate(subject = str_c("A", subject))
```

Use separate to separate one column to multiple columns
```{r}
alien <- alien %>%
  separate(type, into = c("sex", "colour"), sep = "_")
```

Simplify complicated names
```{r}
alien <- alien %>%
  rename(glucose = "glucose - mg/L",
         iq = "IQ - ref to human 100 scale!")
```


Tidy sex column
```{r}
alien <- alien %>%
  mutate(sex = str_sub(sex, start = 1, end = 1))
```

Tidy colour column
```{r}
alien <- alien %>%
  mutate(colour = str_to_lower(colour))
```


# Data exploration

Quick look
```{r}
skim(alien)
```

Try group_by function 
```{r}
alien %>% 
  group_by(colour, region, sex) %>%
  count()
```

Summary statistics
```{r}
alien %>%
  group_by(region) %>%
  summarise(mean_iron = mean(iron, na.rm = TRUE),
            sd_iron = sd(iron, na.rm = TRUE)) 
```

Sort colours in order of popularity
```{r}
alien %>%
  count(colour, sort = TRUE) 
```

Sort by age
```{r}
alien %>%
  arrange(desc(age))
```

Filter dataframe by male green older than 8
```{r}
alien %>%
  filter(sex == "m", age > 8, colour == "green")
```

# Plotting

Scatterplot
```{r}
ggplot(data = alien, aes(x = parse_number(age), y = weight, colour=sex)) + 
  geom_point()
```

Bar chart
```{r}
ggplot(data = alien, aes(x = colour, fill = region)) +
  geom_bar()
```

Histogram
```{r}
ggplot(data = alien, aes(x = glucose, fill = sex)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~sex, ncol = 1) + labs(title="Glucose by sex")
```

Glucose Vs Iron
```{r}
ggplot(data = alien, aes(x = glucose, y = iron)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, linetype = "dotted") +
  facet_wrap(~region) + 
  theme_bw()
```

Recreate some plots

Glucose level by health state
```{r}
ggplot(alien, aes(x = health, y = glucose)) +
  geom_boxplot() +
  labs(x = "Health status",
       y = "Glucose (mg/L)",
       title = "Alien blood glucose by health state",
       subtitle = "March 2060 data")
  
```








# Stats

T test
```{r}
t.test(glucose ~ sex, data = alien)
```

Chi square
```{r}
chisq.test(alien$sex, alien$colour == "purple")
```



Regression
```{r}
mod1 <- lm(iron ~ glucose + region, data=alien)
# model summary
summary(mod1)
# model diagnostics
# plot(mod1)
```

#### HELP


# ggplot2 cheat sheet

```{r}
ggplot(alien, aes(age, glucose, colour = sex)) +
  geom_point() +
  facet_wrap(~ region) +
  labs(x = "Age in years",
       y = "Blood glucose (mg/ml)",
       colour = "Alien sex",
       title = "Glucose vs age by sex and region",
       subtitle = "April 2018") +
  theme_minimal()
```


# Exercise plots

```{r}
ggplot(alien, aes(x = age, y = iron, colour = sex)) +
  geom_point(shape = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 0.2) +
  facet_wrap(~ colour) +
  theme_bw() +
  labs(title = "Iron vs Age by Sex and Colour",
       x = "Age (years)",
       y = "Blood iron (mg/L)",
       colour = "Alien sex")
```

```{r}
ggplot(alien, aes(x = region, y = iron, colour = region)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, shape = 1) +
  theme_bw() +
  labs(title = "Iron vs Region",
       x = "Region",
       y = "Blood iron (mg/L)",
       colour = "Region")
```


```{r}
ggplot(alien, aes(x = iq, fill = region)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ region) +
  labs(title = "IQ vs Region",
       x = "Region",
       y = "IQ score",
       fill = "Region")
```
```{r}
ggplot(alien, aes(x = drives)) +
  geom_bar() +
  labs(title = "What do they drive?",
       x = "Model",
       y = "Count") +
  coord_flip()
```








