---
title: "Alien Study - Example Analysis"
author: "Your name here"
date: "2018-04-04"
output:
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    code_folding: show
---
  
## Load the packages we'll need

```{r, warning=FALSE}
library(tidyverse)
library(readxl)
library(skimr)
library(knitr)
```

## Import data

Load in data

```{r}
alien <- read_excel("aliendata_20180331.xlsx",
                    sheet = 1,
                    na = "-999")
```

## Tidy and clean data

Fill down the missing region data (before bad things happen)
```{r}
alien <- alien %>% 
  fill(region)
```

Remove alien names (it's personal data we don't need!)
```{r}
alien <- alien %>% 
  select(- name)
```

Rename the complicated columns
```{r}
alien <- alien %>% 
  rename(glucose = "glucose - mg/L",
         iq = "IQ - ref to human 100 scale!")
```


Add an 'A' to ID
```{r}
alien <- alien %>% 
  mutate(subject = str_c("A", subject))
```

Separate 'type' into 'sex' and 'colour'
```{r}
alien <- alien %>% 
  separate(type, into = c("sex", "colour"), sep = "_")
```

Use first letter of 'sex' as some are abbreviated
```{r}
alien <- alien %>% 
  mutate(sex = str_sub(sex, 1, 1))
```

Make all the colour lower case as some mixed!
```{r}
alien <- alien %>% 
  mutate(colour = str_to_lower(colour))
```

If age mentions months divide by 12
```{r}
alien <- alien %>% 
  mutate(age = case_when(str_detect(age, "month") ~ parse_number(age) / 12,
                   TRUE ~ parse_number(age)))
```


Data cleaning and tidying over (hopefully!)

## Explore data

An overall summary (uses skim from the skimr package)
```{r}
skim(alien)
```

How many aliens?
```{r}
alien %>% count()
```

How many aliens from each region?
```{r}
alien %>% 
  count(region)
```
How many aliens of each colour in each region?
```{r}
alien %>% 
  count(region, colour)
```
And in order of decreasing frequency...
```{r}
alien %>% 
  count(colour, sort = TRUE)
```
Most common colour
```{r}
alien %>% 
  count(colour, sort = TRUE) %>% 
  top_n(n = 1, wt = n)
```
Most common colour in each region
```{r}
alien %>%
  group_by(region) %>% 
  count(colour, sort = TRUE) %>% 
  top_n(1)
```

### Some summary statistics

mean, median and sd
```{r}
alien %>% 
  summarise(mean(glucose), median(glucose), sd(glucose))
```
summaries for each colour
```{r}
alien %>%
  group_by(colour) %>% 
  summarise(mean = mean(glucose),
            median = median(glucose),
            sd = sd(glucose)) %>% 
  kable(digits = 2, caption = "Glucose (mg/L)")
```


## Visualisation

### Scatter plot

```{r, fig.height=2}
ggplot(alien, aes(x = age, y = weight, colour = sex)) +
  geom_point()
```

### Histogram

```{r}
ggplot(alien, aes(x = glucose)) +
  geom_histogram(binwidth = 2)
```

### Box and whisker plot

```{r}
ggplot(alien, aes(x = region, y = glucose, fill = region)) +
  geom_boxplot()
```

### Jittered plot by colour

```{r}
ggplot(alien, aes(x = colour, y = glucose)) +
  geom_jitter(width = 0.1)
```

### Scatter plot with fitted line

```{r, warning=FALSE}
ggplot(alien, aes(x = age, y = iron)) +
  geom_point() +
  geom_smooth(method = "lm")
```

### Facetting

```{r, warning=FALSE}
ggplot(alien, aes(x = age, y = iron)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ region)
```

### Facetting and colour

```geom_smooth(method = "lm", se = FALSE)``` fits a straight line with no confidence interval)

```{r, warning=FALSE}
ggplot(alien, aes(x = age, y = iron, colour = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ region)
```


### Add some labels and titles, make line thinner andpoints hollow

```{r, warning=FALSE}
ggplot(alien, aes(x = age, y = iron, colour = sex)) +
  geom_point(shape = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 0.2) +
  facet_wrap(~ region) +
  labs(x = "Age (years)",
       y = "Blood Iron (ug/ml)",
       colour = "Sex",
       title = "Alien Iron vs age by sex and region",
       subtitle = "Measured by MS")
```