---
title: "R Workshop Notes"
date: "2018-04-04"
output: 
  tufte::tufte_html:
    tufte_variant: "envisioned"
    highlight: pygments
    css: lab.css
    


---

These are brief notes on data import, tidying, cleaning, exploring and plotting. For much more detail and explanation see Hadley Wickham's book:
`R for Data Science` (R4DS) at (http://r4ds.had.co.nz/wrangle-intro.html).

The import and cleaning code is based on a ficitious data (mydata) so the R code won't run - but the code shows you examples you can modify to suit your datasets and variables.

# Setup

Load useful packages

```{r, eval = FALSE}
library(tidyverse) # all sorts of usefull stuff
library(skimr) # for a summary of your data
library(knitr) # we might use this for tables
```

# Importing data

## From Excel files

Uses use the `read_excel()` function from the `readxl` package. It's installed with the `tidyverse` but you need to load it before using it with...

```{r}
library(readxl)
```

Then to read a sheet called "my-sheet-name-or-number" from an Excel workbook called "my-excel-file.xls", skipping the first 5 rows, into an R object called mydata use... 

```{r, eval = FALSE}
mydata <- read_excel("my-excel-file.xls",
                 sheet = "my-sheet-name-or-number",
                 skip = 5)
```

## From comma-separated value files (csv)

Use the `read_csv()` function from the `readr` package. This is loaded automatically when the `tidyverse` is loaded.

To read a csv file called "my-csv-file.csv" which includes a header row into a new R object called mydata use...

```{r, eval = FALSE}
mydat <- read_csv("my-csv-file.csv")
```

# Tidying data

## Filling down blanks
To fill down any blank cells in the `patient` and `breed` columns of `mydata` use...

```{r, eval = FALSE}
mydata <- mydata %>% 
  fill(patient, breed)
```

## Selecting or removing columns

The `select` verb lets you choose or remove columns. To remove the name column from mydata use...

```{r, eval = FALSE}
mydata <- mydata %>% 
  select(-name)
```

To just select `subject` and all columns between `age` and `glucose` in mydata use...

```{r, eval = FALSE}
mydata <- mydata %>% 
  select(subject, age:glucose)
```

## Renaming columns

To `rename` the column `sex (m/f)` to `sex` use...
```{r, eval = FALSE}
mydata <- mydata %>% 
  rename(sex = `sex (m/f)`)
```
Note that the back ticks `` let you choose columns that have names that would not normally be allowed in R.

## Separating a columns into 2 (or more) new columns

A column that contains data that should be in two separate columns can be split using `separate()`. To split a column called `sex_status` into two columns, `sex` and `status` using `_` as a separator use...

```{r, eval = FALSE}
mydata <- mydata %>% 
  separate(sex_status,
           into = c("sex", "status"),
           sep = "_")
```

This would split "male_neutered" in one column into "male" and "neutered" in two separate columns etc.

## Creating and changing column content

To make a new column use `mutate()`. For example if you have a column `weight` you can add a column log_wt containing `log(weight)` like this...

```{r, eval = FALSE}
mydata <- mydata %>% 
  mutate(log_wt = log(weight))
```

Mutate can also overwrite columns (but be careful). So to convert height in inches to height in cm in the same column you could use...

```{r, eval = FALSE}
mydata <- mydata %>% 
  mutate(height = height * 2.54)
```

## Gathering data in several columns
Sometimes data at different times are stored in one column per time. To convert theses into a column for measurement time and a column for the actual data use the `gather` verb. For example if there were three columns `w_1`, `w_2` and `w_3` all holding weight data at each week these could be converted to a column for week and a column for weight...

```{r, eval = FALSE}
mydata <- mydata %>% 
  gather(key = "week",
         value = "weight",
         w_1, w_2, w_3)
```

This can be tricky at first so for more detail see R4DS (http://r4ds.had.co.nz/tidy-data.html#gathering)



# Exploring  and manipulating data

For the following examples we'll use the `mpg` dataset that comes with the tidyverse. It describes performance data on some cars. The following code chunks should run (as long as the tidyverse and skimr packages are loaded...)

```{r, include = FALSE}
library(tidyverse) # all sorts of usefull stuff
library(skimr) # for a summary of your data
```
```{r, eval = FALSE}
library(tidyverse) # all sorts of usefull stuff
library(skimr) # for a summary of your data
```

## A quick summary

Uses the `skim()` function from the `skimr` package

```{r}
skim(mpg)
```

## Count a text / categorical column

To count / tally the number of cars of each manufacturer

```{r}
mpg %>% 
  count(manufacturer)
```

## Do that again but sort it

```{r}
mpg %>% 
  count(manufacturer, sort = TRUE)
```

## Count a text / categorical column by another one

To count drive system by year (drive systems are forward, rear and 4x4)

```{r}
mpg %>% 
  group_by(year) %>% 
  count(drv)
```

## To calculate the mean and median engine sizes

The `summarise` verb/function reduces a dataframe down to summary columns. The `na.rm = TRUE` argument in the statistics functions tells them to ignore missing values (NA's in R).

```{r}
mpg %>% 
  summarise(mean_displacement = mean(displ, na.rm = TRUE),
            median_displacement = median(displ, na.rm = TRUE))
  
```

We can extend that code to look at engine size by year.

```{r}
mpg %>%
  group_by(year) %>% 
  summarise(mean_displacement = mean(displ, na.rm = TRUE),
            median_displacement = median(displ, na.rm = TRUE))
  
```

Engines look to be getting bigger!

## Filtering observations
To remove (or include) observations that meet a certain criteria use filter. For example to select only the observations from mydata where drive is 4x4
```{r}
mpg %>% 
  filter(drv == "4")
```
Note that to check if a column equals 'something' you use two equals `==`. You can compare numbers using `>` and `<`. To check if a column has values in a  list use `%in%` like this...

```{r}
mpg %>% 
  filter(model %in% c("A4", "civic"))
```

A `!` in front of the test makes it opposite. The following keeps all the observations where the colour **isn't** 'blue'.

```{r}
mpg %>% 
  filter(!model == "a4")
```

Missing values in R are coded as `NA`. There's a special function, `is.na()` to test for missing data. So the following keeps all rows where the colour **isn't** missing

```{r}
mpg %>% 
  filter(!is.na(model))
```


