---
title: "Importing, tidying and cleaning data"
output: 
  html_document:
    highlight: pygments
    css: lab.css
---

```{r, include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

These are brief notes on data import, tidying and cleaning. For much more detail and explanation see Hadley Wickham's book:
`R for Data Science` (R4DS) at (http://r4ds.had.co.nz/wrangle-intro.html).

# Load the tidyverse

Before importing or tidying data run...
```{r}
library(tidyverse)
```
...to load the `tidyverse` set of packages first.

# Importing data

## From Excel files

Uses use the `read_excel()` function from the `readxl` package. It's installed with the `tidyverse` but you need to load it before using it with...

```{r}
library(readxl)
```

Then to read a sheet called "my-sheet-name-or-number" from an Excel workbook called "my-excel-file.xls", skipping the first 5 rows, into an R object called mydata use... 

```{r}
mydata <- read_excel("my-excel-file.xls",
                 sheet = "my-sheet-name-or-number",
                 skip = 5)
```

## From comma-separated value files (csv)

Use the `read_csv()` function from the `readr` package. This is loaded automatically when the `tidyverse` is loaded.

To read a csv file called "my-csv-file.csv" which includes a header row into a new R object called mydata use...

```{r}
mydat <- read_csv("my-csv-file.csv")
```

# Tidying data

## Filling down blanks
To fill down any blank cells in the `patient` and `breed` columns of `mydata` use...

```{r}
mydata <- mydata %>% 
  fill(patient, breed)
```

## Selecting or removing columns

The `select` verb lets you choose or remove columns. To remove the name column from mydata use...

```{r}
mydata <- mydata %>% 
  select(-name)
```

To just select `subject` and all columns between `age` and `glucose` in mydata use...

```{r}
mydata <- mydata %>% 
  select(subject, age:glucose)
```

## Renaming columns

To `rename` the column `sex (m/f)` to `sex` use...
```{r}
mydata <- mydata %>% 
  rename(sex = `sex (m/f)`)
```
Note that the backticks `` let you choose columns that have names that would not normally be allowed in R.

## Separating a columns into 2 (or more) new columns

A column that contains data that should be in two separate columns can be split using `separate()`. To split a column called `sex_status` into two columns, `sex` and `status` using `_` as a separator use...

```{r}
mydata <- mydata %>% 
  separate(sex_status,
           into = c("sex", "status"),
           sep = "_")
```

This would split "male_neutered" in one column into "male" and "neutered" in two separate columns etc.

## Creating and changing column content

To make a new column use `mutate()`. For example if you have a column `weight` you can add a column log_wt containing `log(weight)` like this...

```{r}
mydata <- mydata %>% 
  mutate(log_wt = log(weight))
```

Mutate can also overwrite columns (but be carefule). So to convert height in inches to height in cm in the same column you could use...

```{r}
mydata <- mydata %>% 
  mutate(height = height * 2.54)
```

## Gathering data in several columns
Sometimes data at diffrent times are stored in one column per time. To convert theses into a column for meaurement time and a column for the actual data use the `gather` verb. For example if there were three columns `w_1`, `w_2` and `w_3` all holding weight data at each week these could be converted to a column for week and a column for weight...

```{r}
mydata <- mydata %>% 
  gather(key = "week",
         value = "weight",
         w_1, w_2, w_3)
```

This can be tricky at first so for more detail see R4DS (http://r4ds.had.co.nz/tidy-data.html#gathering)








