---
title: "Alien analysis & workshop notes"
author: "Ian Handel"
date: "11/04/2018"
output: 
  html_document: 
    code_folding: show
    highlight: pygments
    theme: cerulean
    toc: yes
    toc_float: yes
---
## Introduction

Here's the tidied up notes from the 'alien-analysis session'. This document was made by 'knitting' an RMarkdown file which loads in the alien data from an Excel workbook (in the same folder), cleans it and then analyses it. The RMarkdown file is very similar to the one you will have made in the workshop. We've added some lines to the header to make a table of contents and to change the colour scheme. There's also more text added to explain what's going on. Finally we've added some further code to demonstrate more features of R including data management, plotting and some basic statistics. When you're reading through these notes you can also look at the ```alien_analysis_20180404.Rmd``` RMarkdown file to see the code that made it.

We'll include web links to Hadley Wickham and Garrett Grolemund's excellent R for Data Science book [R4DS](http://r4ds.had.co.nz) and other sources with more detail where relevant.

R code is shown in these notes in boxes with a light grey background. R output is shown in boxes with a white background with '##' at the left hand side of each line.


## Load Packages 

First job is to load the add-in packages that we'll need. When you load packages there'll normally be quite a few warnings and messages.

```{r}
#load libraries
library(tidyverse)
library(readxl)
library(skimr)
library(knitr)
```

## Get data in

As the data is stored in Excel we use the ```read_excel()``` function from the `readxl` package to import it. This function is given the filename, the sheet name or number (```sheet=```) and optionally the codes for missing data (```na=```). Our code reads the first sheet from "alien_analysis_20180404.Rmd" into a new R object called `alien`. Remember that we're not changing the Excel sheet - just reading it into R and putting it into an object we're calling `alien`.

```{r}
alien <- read_excel("aliendata_20180404.xlsx", sheet = 1, na = "-999")
```

Once the data is in R we can look at the top of it by clicking on its name in the "environment" tab/panel (top right part of the RSTudio window) or by typing ```View(alien)``` into the console (note that "View"" has a capital "V"). Once we've had a look we can start cleaning and tidying it ready for analysis.

## Data tidy and clean

### Fill missing regions
First thing to do is to fill down the missing cells in the "region" column. The following R code takes the ```alien``` object, pipes it into the ```fill``` function/verb and fills the "region" column. The result is put back into the ```alien``` object with ```<-```. Remember, this changes the ```alien``` object in R but doesn't alter the spreadsheet. This is important. It means our original data is untouched and is always available if we change out minds about our tidying!

If you want to look at what the code is doing one step at a time you can run each code chunk in the ```alien_analysis_20180404.Rmd``` file at a time (using the little green arrow at its right hand side) and keep looking back at the ```alien``` data to see it change.  

```{r}
alien <- alien %>%
  fill(region)
```

### Remove names

Next we'll remove the alien names - it's personal data that we don't need and would rather not have on our screens (aliens have rights too!). The ```select()``` verb/function lets us do this. If we give it column names either one at a time or in a range with ':' between them it will ***keep*** those columns. If we put a "-" infront of one or more column names it will remove them. See [R4DS](http://r4ds.had.co.nz/transform.html#select-columns-with-select) for more detailed instructions on using ```select()```.

```{r}
alien <- alien %>%
  select(-name)
```

### Improve the subject iD format
Teh subject ID is currebtly a number. Wwe prefer ID's to be formatted so thay are unambiguously an 'ID' and to be the same number of charcters long. We use a new verb/function here `mutate`. You pipe a data frame into mutate and give it the name of the column you want to create then `=` and then an expression (recispe) for the column. If you create an existing column it effectively overwirites it. So first off we overwrite subject using the str_pad function (this function is in the "stringr")

Tidy ID column
```{r}
alien <- alien %>%
  mutate(subject = str_pad(subject, width = 3, side = "left", pad = "0")) %>% 
  mutate(subject = str_c("A", subject))
```

Separate type column 
```{r}
alien <- alien %>%
  separate(type, into = c("sex", "colour"), sep = "_")
```

Take the first letter of sex column and correct colour column
```{r}
library(stringr)
alien <- alien %>%
  mutate(sex = str_sub(sex, start = 1, end = 1)) %>%
  mutate(colour = str_to_lower(colour))
```

Rename columns
```{r}
alien <- alien %>%
  rename(glucose = "glucose - mg/L",
         iq = "IQ - ref to human 100 scale!")
```

## Data exploration
Quick look
```{r}
skim(alien)
```

Tabulations
```{r}
alien %>%
  count(colour, sex)
```

Sort colours in order of popularity
```{r}
alien %>%
  count(colour, sort = TRUE) 
```

Sort by age
```{r}
alien %>%
  arrange(desc(age))
```

Filter dataframe by male green older than 8
```{r}
alien %>%
  filter(sex == "m", age > 8, colour == "green")
```

Summarise numerical columns
```{r}
alien %>%
  group_by(colour, sex) %>%
  summarise(mean_glucose = mean(glucose),
            sd_glucose = sd(glucose))
```

Summary statistics
```{r}
alien %>%
  group_by(region) %>%
  summarise(mean_iron = mean(iron, na.rm = TRUE),
            sd_iron = sd(iron, na.rm = TRUE)) 
```


## Plotting!!!!

Scatterplot
```{r}
ggplot(data = alien, aes(x = age, y = weight, colour = sex)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~region)
```

Boxplot
```{r}
ggplot(alien, aes(x= region, y = glucose, colour = sex)) +
  geom_boxplot() + 
  geom_jitter(width=0.2, size=0.5) + 
  theme_bw()
```

Histogram 
```{r}
ggplot(alien, aes(x = glucose, fill = sex)) + 
  geom_histogram(bins = 50) + 
  facet_wrap(~sex, ncol = 1)
```

Glucose level by health state
```{r}
ggplot(alien, aes(x = health, y = glucose)) +
  geom_boxplot() +
  labs(x = "Health status",
       y = "Glucose (mg/L)",
       title = "Alien blood glucose by health state",
       subtitle = "March 2060 data")
  
```


```{r}
ggplot(alien, aes(x = age, y = iron, colour = sex)) +
  geom_point(shape = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 0.2) +
  facet_wrap(~ colour) +
  theme_bw() +
  labs(title = "Iron vs Age by Sex and Colour",
       x = "Age (years)",
       y = "Blood iron (mg/L)",
       colour = "Alien sex")
```

```{r}
ggplot(alien, aes(x = region, y = iron, colour = region)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, shape = 1) +
  theme_bw() +
  labs(title = "Iron vs Region",
       x = "Region",
       y = "Blood iron (mg/L)",
       colour = "Region")
```


```{r}
ggplot(alien, aes(x = iq, fill = region)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ region) +
  labs(title = "IQ vs Region",
       x = "Region",
       y = "IQ score",
       fill = "Region")
```
```{r}
ggplot(alien, aes(x = drives)) +
  geom_bar() +
  labs(title = "What do they drive?",
       x = "Model",
       y = "Count") +
  coord_flip()
```


## Stats

T test
```{r}
t.test(glucose ~ sex, data = alien)
```

Chi square
```{r}
chisq.test(alien$sex, alien$colour == "purple")
```

## Regression
```{r}
mod1 <- lm(iron ~ glucose + region, data=alien)
# model summary
summary(mod1)
# model diagnostics
# plot(mod1)
```








